\thetitle

### 2020-10-23 sun_phases/spin_qudit_tomo.tex
added note concerning a tightened error bound

\newcommand{\B}{\mathcal{B}}
\newcommand{\N}{\mathcal{N}}
\renewcommand{\O}{\mathcal{O}}
\newcommand{\Q}{\mathcal{Q}}
\newcommand{\R}{\mathcal{R}}
\newcommand{\T}{\mathcal{T}}
\newcommand{\Y}{\mathcal{Y}}

@@

A set of measurement axes $V=\set{\v v}$ nominally induces a set of projectors $\v\Pi_V=\set{\Pi_j^V}$ that will be measured in an experiment, where for shorthand we use a combined index $j\simeq\p{\v v_j,\mu_j}$ that specifies both a measurement axis $\v v_j$ and an outcome $\mu_j$, and for clarify we index objects by $V$ to emphasize their dependence on the choice of measurement axes.
The projectors $\v\Pi_V$ can be used to construct the overlap (Gram) matrix
\begin{align}
  G_V = \sum_{i,j=1}^{\abs{V}\times d} \obk{\Pi_i^V|\Pi_j^V} \op{i}{j}.
  \label{eq:overlap_matrix}
\end{align}
Eigenvectors of $G_V$ define linearly independent linear combinations of the projectors in $\v\Pi_V$.
If the projectors in $\v\Pi_V$ span the entire space of operators on a $d$-level spin qudit, which is a necessary and sufficient condition for $V$ to allow for full state tomography, then $G_V$ must have exactly $d^2$ nonzero eigenvalues.
Indexing these eigenvalues $G^V_k$ and the corresponding (normalized) eigenvectors $\v x^V_k\equiv(x^V_{k,1},x^V_{k,2},\cdots)$ by an integer $k\in\set{1,2,\cdots,d^2}$,
we can construct the orthonormal operators
\begin{align}
  Q^V_k \equiv \sum_j g^V_{kj} \Pi_j^V,
  &&
  g^V_{kj} \equiv \f{x^V_{kj}}{\sqrt{G^V_k}}.
\end{align}
These operators allow us to expand any state $\rho$ of a $d$-level spin qudit in the form
\begin{align}
  \rho = \sum_{k=1}^{d^2} \rho_k^V Q_k^V,
  &&
  \rho_k^V \equiv \bk{Q^V_k}_\rho.
\end{align}
Given empirical estimates $E_j^V$ of the expectation values $\bk{\Pi_j^V}_\rho$, an empirical estimate $\tilde\rho$ of $\rho$ is then
\begin{align}
  \tilde\rho \equiv \sum_k \tilde\rho^V_k Q^V_k,
  \label{eq:reconstructed_state}
\end{align}
where
\begin{align}
  \tilde\rho^V_k \equiv \sum_j g^V_{kj} E_j^V
  \approx \sum_j g^V_{kj} \bk{\Pi_j^V}_\rho
  = \bk{Q^V_k}_\rho
  = \rho^V_k.
\end{align}
The eigenvalues and eigenvectors of the overlap matrix $G_V$ provide strict upper bounds on the distance between the empirical estimate $\tilde\rho$ and the true state $\rho$.
If the estimates $E^V_j$ approximate $\bk{\Pi^V_j}_\rho$ to within an error $\epsilon$, i.e.
\begin{align}
  \abs{E^V_j-\bk{\Pi^V_j}_\rho} \le \epsilon,
\end{align}
then the error in the estimate $\tilde\rho^V_k$ of $\rho^V_k$ is
\begin{align}
  \abs{\tilde\rho^V_k-\rho^V_k}
  &= \abs{\sum_j g_{kj}^V \p{E^V_j-\bk{\Pi^V_j}_\rho}} \\
  &\le \sum_j \abs{g_{kj}^V \p{E^V_j-\bk{\Pi^V_j}_\rho}} \\
  &\le \epsilon \norm{\v g^V_k}_1,
\end{align}
where $\norm{\v g}_1=\sum_j\abs{g_j}$ is the 1-norm of the vector $\v g$.
Using the fact that the operators $Q^V_k$ are orthonormal, we can bound the 2-norm (Euclidean) distance $\norm{\tilde\rho-\rho}_2$ between $\tilde\rho$ and $\rho$ as
\begin{align}
  \norm{\tilde\rho-\rho}_2^2
  = \sum_k \abs{\tilde\rho^V_k - \rho^V_k}^2
  \le \epsilon^2 \N_V^2,
  \label{eq:bound}
\end{align}
where
\begin{align}
  \N_V^2 \equiv \min_U \sum_k \norm{U\v g^V_k}_1^2
  = \min_U \sum_k \f{\norm{U\v x^V_k}_1^2}{G^V_k}
  \label{eq:norm}
\end{align}
is a norm uniquely determined by the set of measurement axes $V$, with minimization performed over unitary transformations $U$ of all $\v g^V_k$ and $\v x^V_k$ to account for the fact that their 1-norms are generally basis-dependent.
Seen otherwise, this minimization accounts for the fact that we can perform the analysis between Eqs.~\eqref{eq:overlap_matrix} and \eqref{eq:bound} identically with any $G_V\to U G_V U^\dag$, where $U$ is an arbitrary unitary that mixes the projectors in $\v\Pi_V$.

To better understand the bound in Eq.~\eqref{eq:bound}, we observe that by a change of operator basis, or equivalently a particular choice of unitary $U=U_T$, measuring the projectors $\set{\Pi_{\v v\mu}}$ is equivalent to measuring the transition operators $\set{T_{\v v\ell,0}}$.
Orthogonality of transition operators with different degrees $\ell$ then implies that the overlap matrix
\begin{align}
  G_V^{\t{trn}} \equiv U_T G_V U_T^\dag
  = \sum_{\ell,\v v,\v v'} \obk{T_{\v v\ell,0}|T_{\v v'\ell,0}}
  \op{\v v\ell}{\v v'\ell}
\end{align}
has a block-diagonal structure, with $d$ blocks of size $\abs{V}\times\abs{V}$.
Eigenvectors of $G_V$ can therefore have at most $\abs{V}$ nonzero components in the transition operator basis, which implies that
\begin{align}
  \min_U \norm{U\v x^V_k}_1^2
  \le \norm{U_T\v x^V_k}_1^2
  \le \abs{V}\, \norm{\v x^V_k}_2^2
  = \abs{V},
\end{align}
so
\begin{align}
  \N_V^2 \le \abs{V} \sum_k \f1{G^V_k}.
\end{align}
Note that this sum, as well as that in Eq.~\eqref{eq:norm}, is performed over the $d^2$ largest eigenvalues of $G_V$, and therefore may be divergent if $G_V$ has fewer than $d^2$ nonzero eigenvalues.
Altogether, diagonalizing each of $d$ blocks of $G_V$ takes a time that is $O(\abs{V}^3)$, so an upper bound for $\N_V$ can be computed in $O(d\abs{V}^3)$ serial or $O(\abs{V}^3)$ parallel runtime.

The bound in Eq.~\eqref{eq:bound} motivates the following simple protocol for spin qudit tomography: select a random set of measurement axes $V$ by uniformly sampling points on the sphere \footnote{To sample a point $\p{\alpha,\beta}$ from the uniform distribution on the sphere (with azimuthal angle $\alpha$ and polar angle $\beta$), you can sample a point $\p{a,b}\in[0,1]\times[0,1]$ from the uniform distribution on the unit square, then set $\alpha=2\pi a$ and $\beta=\arccos\p{1-2b}$.}, and variationally optimize the $2\abs{V}$ parameters in $V$ (two angles for each point $\v v\in V$) to minimize the cost function
\begin{align}
  \tilde\N_V^2 \equiv \sum_k \f{\norm{U_T\v x^V_k}_1^2}{G^V_k}
  \ge \N_V^2,
\end{align}
where $U_T\v x^V_k$ are (normalized) eigenvectors of the overlap matrix $G_V$ in the transition operator basis, obtained by diagonalizing $G_V^{\t{trn}}$.
If $\abs{V}$ is too large for variational optimization to be practical, one can instead generate many random sets of measurement axes, and then choose the set that minimizes $\tilde\N_V^2$.
Upon choosing a set of measurement axes $V$, experimentally characterize the expectation values $\bk{\Pi^V_j}_\rho$.
If these measurements can approximate the expectation values $\bk{\Pi^V_j}_\rho$ to within a known error $\epsilon$, then the empirical estimate $\tilde\rho$ defined in Eq.~\eqref{eq:reconstructed_state} is an approximation of the true state $\rho$ to within a 2-norm distance of $\epsilon\N_V$.
If the empirical estimate $\tilde\rho$ has negative eigenvalues, its 2-norm distance from $\rho$ can be further reduced via maximum-likelihood corrections \cite{smolin2012efficient}.
In terms of classical computing costs before and after the estimation of expectation values $\bk{\Pi^V_j}_\rho$, this tomography protocol is bottlenecked by the need to diagonalize the overlap matrix $G_V^{\t{trn}}$, which has $d$ blocks of size $\abs{V}\times\abs{V}$.
Altogether, this protocol has an $O(d\abs{V}^3)$ serial or $O(\abs{V}^3)$ parallel runtime.

>>>>

\newcommand{\B}{\mathcal{B}}
\newcommand{\M}{\mathcal{M}}
\newcommand{\N}{\mathcal{N}}
\renewcommand{\O}{\mathcal{O}}
\newcommand{\Q}{\mathcal{Q}}
\newcommand{\R}{\mathcal{R}}
\newcommand{\T}{\mathcal{T}}
\newcommand{\Y}{\mathcal{Y}}

@@

A set of measurement axes $V=\set{\v v}$ nominally induces a set of projectors $\v\Pi_V=\set{\Pi_j^V}$ that will be measured in an experiment, where for shorthand we use a combined index $j\simeq\p{\v v_j,\mu_j}$ that specifies both a measurement axis $\v v_j$ and an outcome $\mu_j$, and for clarify we index objects by $V$ to emphasize their dependence on the choice of measurement axes.
The projectors $\v\Pi_V$ can be used to construct the overlap (Gram) matrix
\begin{align}
  G_V = \sum_{i,j=1}^{\abs{V}\times d} \obk{\Pi_i^V|\Pi_j^V} \op{i}{j}.
  \label{eq:overlap_matrix}
\end{align}
Eigenvectors of $G_V$ define linearly independent linear combinations of the projectors in $\v\Pi_V$.
If the projectors in $\v\Pi_V$ span the entire space of operators on a $d$-level spin qudit, which is a necessary and sufficient condition for $V$ to allow for full state tomography, then $G_V$ must have exactly $d^2$ nonzero eigenvalues.
Indexing these eigenvalues $G^V_k$ and the corresponding (normalized) eigenvectors $\v x^V_k\equiv(x^V_{k,1},x^V_{k,2},\cdots)$ by an integer $k\in\set{1,2,\cdots,d^2}$,
we can construct the orthonormal operators
\begin{align}
  Q^V_k \equiv \sum_j g^V_{kj} \Pi_j^V,
  &&
  g^V_{kj} \equiv \f{x^V_{kj}}{\sqrt{G^V_k}}.
\end{align}
These operators allow us to expand any state $\rho$ of a $d$-level spin qudit in the form
\begin{align}
  \rho = \sum_{k=1}^{d^2} \rho_k^V Q_k^V,
  &&
  \rho_k^V \equiv \bk{Q^V_k}_\rho.
\end{align}
Given empirical estimates $E_j^V$ of the expectation values $\bk{\Pi_j^V}_\rho$, an empirical estimate $\tilde\rho$ of $\rho$ is then
\begin{align}
  \tilde\rho \equiv \sum_k \tilde\rho^V_k Q^V_k,
  \label{eq:reconstructed_state}
\end{align}
where
\begin{align}
  \tilde\rho^V_k \equiv \sum_j g^V_{kj} E_j^V
  \approx \sum_j g^V_{kj} \bk{\Pi_j^V}_\rho
  = \bk{Q^V_k}_\rho
  = \rho^V_k.
\end{align}
The eigenvalues and eigenvectors of the overlap matrix $G_V$ provide strict upper bounds on the distance between the empirical estimate $\tilde\rho$ and the true state $\rho$.
If the estimates $E^V_j$ approximate $\bk{\Pi^V_j}_\rho$ to within an error $\epsilon$, i.e.
\begin{align}
  \abs{E^V_j-\bk{\Pi^V_j}_\rho} \le \epsilon,
\end{align}
then the error in the estimate $\tilde\rho^V_k$ of $\rho^V_k$ is
\begin{align}
  \abs{\tilde\rho^V_k-\rho^V_k}
  &= \abs{\sum_j g_{kj}^V \p{E^V_j-\bk{\Pi^V_j}_\rho}} \\
  &\le \sum_j \abs{g_{kj}^V \p{E^V_j-\bk{\Pi^V_j}_\rho}} \\
  &\le \epsilon \norm{\v g^V_k}_1,
\end{align}
where $\norm{\v g}_1=\sum_j\abs{g_j}$ is the 1-norm of the vector $\v g$.
Using the fact that the operators $Q^V_k$ are orthonormal, we can bound the 2-norm (Euclidean) distance $\norm{\tilde\rho-\rho}_2$ between $\tilde\rho$ and $\rho$ as
\begin{align}
  \norm{\tilde\rho-\rho}_2^2
  = \sum_k \abs{\tilde\rho^V_k - \rho^V_k}^2
  \le \epsilon^2 \N_V^2,
  \label{eq:bound}
\end{align}
where
\begin{align}
  \N_V^2 \equiv \min_U \sum_k \norm{U\v g^V_k}_1^2
  = \min_U \sum_k \f{\norm{U\v x^V_k}_1^2}{G^V_k}
  \label{eq:norm}
\end{align}
is a norm uniquely determined by the set of measurement axes $V$, with minimization performed over unitary transformations $U$ of all $\v g^V_k$ and $\v x^V_k$ to account for the fact that their 1-norms are generally basis-dependent.
Seen otherwise, this minimization accounts for the fact that we can repeat the analysis between Eqs.~\eqref{eq:overlap_matrix} and \eqref{eq:bound} with any $G_V\to U G_V U^\dag$, where $U$ is an arbitrary unitary that mixes the projectors in $\v\Pi_V$.

To better understand the bound in Eq.~\eqref{eq:bound}, we observe that by a change of operator basis, or equivalently a particular choice of unitary $U_T$, measuring the projectors $\set{\Pi_{\v v\mu}}$ is equivalent to measuring the transition operators $\set{T_{\v v\ell,0}}$.
Orthogonality of transition operators with different degrees $\ell$ then implies that the overlap matrix
\begin{align}
  G_V^{\t{trn}} \equiv U_T G_V U_T^\dag
  = \sum_{\ell,\v v,\v v'} \obk{T_{\v v\ell,0}|T_{\v v'\ell,0}}
  \op{\v v\ell}{\v v'\ell}
\end{align}
has a block-diagonal structure, with $d$ blocks of size $\abs{V}\times\abs{V}$.
Eigenvectors of $G_V$ can therefore have at most $\abs{V}$ nonzero components in the transition operator basis, which implies that
\begin{align}
  \N_V \le \sum_k \f{\norm{U_T\v x^V_k}_1^2}{G^V_k}
  \le \sum_k \f{\abs{V}\,\norm{U_T\v x^V_k}_2^2}{G^V_k}
  = \sum_k \f{\abs{V}}{G^V_k}.
\end{align}
Note that these sums are performed over the $d^2$ largest eigenvalues of $G_V$ (or $G_V^{\t{trn}}$), and therefore may be divergent if $G_V$ has fewer than $d^2$ nonzero eigenvalues.
Altogether, diagonalizing each of $d$ blocks of $G_V^{\t{trn}}$ takes a time that is $O(\abs{V}^3)$, so an upper bound for $\N_V$ can be computed in $O(d\abs{V}^3)$ serial or $O(\abs{V}^3)$ parallel runtime.
[\red{Note: rather than an absolute error $\epsilon$ on experimental measurements, we can assume that $E_j$ is equal to $\bk{\Pi_j}_\rho$ up to Gaussian noise with variance $\epsilon^2$.  In that case, rather than a strict error bound we can make statements about the expected (squared) distance between $\tilde\rho$ and $\rho$, which turns out to be simply $\epsilon^2\sum_k1/G^V_k$.  In retrospect, I think that this is a much nicer result, but I haven't written it up yet, which will require me to restructure this section somewhat.}]

The bound in Eq.~\eqref{eq:bound} motivates the following simple protocol for spin qudit tomography: select a random set of measurement axes $V$ by uniformly sampling points on the sphere \footnote{To sample a point $\p{\alpha,\beta}$ from the uniform distribution on the sphere (with azimuthal angle $\alpha$ and polar angle $\beta$), you can sample a point $\p{a,b}\in[0,1]\times[0,1]$ from the uniform distribution on the unit square, then set $\alpha=2\pi a$ and $\beta=\arccos\p{1-2b}$.}, and variationally optimize the $2\abs{V}$ parameters in $V$ (two angles for each point $\v v\in V$) to minimize the cost function
\begin{align}
  \tilde\N_V^2 \equiv \sum_k \f{\norm{\v y^V_k}_1^2}{G^V_k}
  \ge \N_V^2,
\end{align}
where $\v y^V_k$ are the (normalized) eigenvectors of the overlap matrix $G_V^{\t{trn}}$ in the transition operator basis, with corresponding eigenvalues $G^V_k$.
If $\abs{V}$ is too large for variational optimization to be practical, one can instead generate many random sets of measurement axes, and then choose the set that minimizes $\tilde\N_V^2$.
Upon choosing a set of measurement axes $V$, experimentally characterize the expectation values $\bk{\Pi^V_j}_\rho$.
If these measurements can approximate the expectation values $\bk{\Pi^V_j}_\rho$ to within a known error $\epsilon$, then the empirical estimate $\tilde\rho$ defined in Eq.~\eqref{eq:reconstructed_state} is an approximation of the true state $\rho$ to within a 2-norm distance of $\epsilon\N_V$.
If the empirical estimate $\tilde\rho$ has negative eigenvalues, its 2-norm distance from $\rho$ can be further reduced via maximum-likelihood corrections \cite{smolin2012efficient}.
In terms of classical computing costs before and after the estimation of expectation values $\bk{\Pi^V_j}_\rho$, this tomography protocol is bottlenecked by the need to diagonalize the overlap matrix $G_V^{\t{trn}}$, which has $d$ blocks of size $\abs{V}\times\abs{V}$.
Altogether, this protocol has an $O(d\abs{V}^3)$ serial or $O(\abs{V}^3)$ parallel runtime.
